---
title: "baltree_unequal_popsizes_binary_prior"
author: "Annie Xie"
date: "2025-12-08"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction

In this analysis, I investigate symEBcovMF in the balanced tree setting with unequal population sizes. I've found that other methods do not perform as well in this setting, so I want to see how symEBcovMF does.

# Packages and Functions

```{r}
devtools::load_all('~/Documents/Github/ebnm')
library(pheatmap)
library(ggplot2)
library(symebcovmf)
```

```{r}
source('code/visualization_functions.R')
```

```{r}
compute_L2_fit <- function(est, dat){
  score <- sum((dat - est)^2) - sum((diag(dat) - diag(est))^2)
  return(score)
}
```

```{r}
compute_crossprod_similarity <- function(est, truth){
  K_est <- ncol(est)
  K_truth <- ncol(truth)
  n <- nrow(est)
  
  #if estimates don't have same number of columns, try padding the estimate with zeros and make cosine similarity zero
  if (K_est < K_truth){
    est <- cbind(est, matrix(rep(0, n*(K_truth-K_est)), nrow = n))
  }
  
  if (K_est > K_truth){
    truth <- cbind(truth, matrix(rep(0, n*(K_est - K_truth)), nrow = n))
  }
  
  #normalize est and truth
  norms_est <- apply(est, 2, function(x){sqrt(sum(x^2))})
  norms_est[norms_est == 0] <- Inf
  
  norms_truth <- apply(truth, 2, function(x){sqrt(sum(x^2))})
  norms_truth[norms_truth == 0] <- Inf
  
  est_normalized <- t(t(est)/norms_est)
  truth_normalized <- t(t(truth)/norms_truth)
  
  #compute matrix of cosine similarities
  cosine_sim_matrix <- abs(crossprod(est_normalized, truth_normalized))
  
  assignment_problem <- lpSolve::lp.assign(t(cosine_sim_matrix), direction = "max")
  return((1/K_truth)*assignment_problem$objval)
}
```

# Data Generation

```{r}
sim_4pops <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(7 * p, sd = rep(args$branch_sds, each = p)), ncol = 7)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  LL <- matrix(0, nrow = n, ncol = 7)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 1, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 0, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(1, 0, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 1, 0, 0), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 1, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args = list(pop_sizes = c(20,50,30,60), n_genes = 1000, branch_sds = rep(2,7), indiv_sd = 1, seed = 1)
sim_data <- sim_4pops(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- c(rep('A', 20), rep('B', 50), rep('C', 30), rep('D', 60))
plot_loadings(sim_data$LL, pop_vec)
```

This is a plot of the eigenvalues of the Gram matrix:
```{r}
S_eigen <- eigen(sim_data$YYt)
plot(S_eigen$values) + abline(a = 0, b = 0, col = 'red')
```

This is the minimum eigenvalue:
```{r}
min(S_eigen$values)
```

# symEBcovMF with binary prior

First, we start with running greedy symEBcovMF with the generalized-binary prior. I set the scale to make the prior more binary.
```{r}
ebnm_generalized_binary_fix_scale <- function(x, s, mode = 'estimate', g_init = NULL, fix_g = FALSE, output = ebnm_output_default(), control = NULL){
  ebnm_gb_output <- ebnm::ebnm_generalized_binary(x = x, s = s, mode = mode, 
                                                  scale = 0.01, 
                                                  g_init = g_init, fix_g = fix_g,
                                                  output = output, control = control)
  return(ebnm_gb_output)
}
```

```{r, results = 'hide', warning = FALSE}
symebcovmf_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_generalized_binary_fix_scale, K = 7, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}_{pt-exp}$, the estimate from symEBcovMF:
```{r}
plot_loadings(symebcovmf_fit$L_pm %*% diag(sqrt(symebcovmf_fit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_fit$elbo
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit$L_pm, sim_data$LL)
```

## Observations
The greedy method is able to find something that looks like the tree.

## Additional Backfit

Now, we run additional backfit. This is the code for the backfit:
```{r, results = 'hide', warning = FALSE}
symebcovmf_fit_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_fit, ebnm_fn = ebnm_generalized_binary_fix_scale, backfit_maxiter = 500)
```

This is a scatter plot of $\hat{L}_{pt-exp-backfit}$, the estimate from symEBcovMF with backfit:
```{r}
plot_loadings(symebcovmf_fit_backfit$L_pm %*% diag(sqrt(symebcovmf_fit_backfit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_fit_backfit$elbo
```

This is a plot of the progression of the ELBO:
```{r}
plot(symebcovmf_fit_backfit$backfit_iter_elbo_vec, ylab = 'ELBO')
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit_backfit$L_pm, sim_data$LL)
```

### Observations

The final answer after the backfit still looks like a tree.

# symEBcovMF with more flexible binary prior

Now, we will try the generalized-binary prior with a different scale. I set the scale to make the prior less strictly binary.

```{r}
symebcovmf_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm::ebnm_generalized_binary, K = 7, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}_{pt-exp}$, the estimate from symEBcovMF:
```{r}
plot_loadings(symebcovmf_fit$L_pm %*% diag(sqrt(symebcovmf_fit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_fit$elbo
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit$L_pm, sim_data$LL)
```

## Observations
With the less binary prior, the method is able to find some of the branches. However, for a couple of the branches, it finds groupings of two populations rather than a single population. Furthermore, some of the branch factors are not very binary.

## Additional Backfit

Now, we run additional backfit. This is the code for the backfit:
```{r, results = 'hide', warning = FALSE}
symebcovmf_fit_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_fit, ebnm_fn = ebnm::ebnm_generalized_binary, backfit_maxiter = 500)
```

This is a scatter plot of $\hat{L}_{pt-exp-backfit}$, the estimate from symEBcovMF with backfit:
```{r}
plot_loadings(symebcovmf_fit_backfit$L_pm %*% diag(sqrt(symebcovmf_fit_backfit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_fit_backfit$elbo
```

This is a plot of the progression of the ELBO:
```{r}
plot(symebcovmf_fit_backfit$backfit_iter_elbo_vec, ylab = 'ELBO')
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit_backfit$L_pm, sim_data$LL)
```

### Observations

With additional backfitting, the final answer does not look closer to a tree.

## Initialize with true values

```{r}
symebcovmf_true_value_init <- sym_ebcovmf_init(sim_data$YYt)

init_L_norms <- apply(sim_data$LL, 2, function(x){sqrt(sum(x^2))})
init_L_normalized <- t(t(sim_data$LL)/init_L_norms)
init_lambda <- (init_L_norms^2)*diag(crossprod(sim_data$FF)/ncol(sim_data$Y))
  
symebcovmf_true_value_init <- sym_ebcovmf_factors_init(sim_data$YYt, symebcovmf_true_value_init, init_L_normalized, init_lambda)
```

```{r}
symebcovmf_true_value_init <- refit_lambda(S = sim_data$YYt, 
                                 symebcovmf_true_value_init, 
                                 maxiter = 20)
```

```{r, results = 'hide', warning = FALSE}
symebcovmf_true_value_init <- sym_ebcovmf_backfit(sim_data$YYt, 
                                        symebcovmf_true_value_init, 
                                        ebnm_fn = ebnm::ebnm_generalized_binary, 
                                        backfit_maxiter = 500)
```

This is a scatter plot of $\hat{L}_{pt-exp-backfit}$, the estimate from symEBcovMF with backfit:
```{r}
plot_loadings(symebcovmf_true_value_init$L_pm %*% diag(sqrt(symebcovmf_true_value_init$lambda)), pop_vec)
```

```{r}
symebcovmf_true_value_init$elbo
```

### Observations

This fit has a higher objective function than the fit with the more flexible generalized binary. This suggests that the method prefers the binary tree structure. Perhaps with a longer backfit, the fit initialized with the greedy fit would look closer to the binary tree representation.
