---
title: "pt_laplace_split_init_symebcovmf_exploration"
author: "Annie Xie"
date: "2025-11-20"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction

In this analysis, I explore the performance of symEBcovMF with the point-laplace plus splitting initialization.

## Motivation
The reason I am exploring this initialization is because the GBCD method uses this point-Laplace prior fit plus splitting initialization procedure. This initialization seems to contribute to GBCD's performance, so I want to see how the procedure performs with symEBcovMF. Since the initialization works well with GBCD, I expect that it will also work well with symEBcovMF. There's no reason to believe the additional symmetry constraint should affect the performance of the initialization. (Note I am fitting the point-Laplace fit with flashier, not with symEBcovMF. But I imagine the procedure would still work well if you used symEBcovMF to get the point-Laplace prior fit.)

# Packages and Functions

```{r}
devtools::load_all('~/Documents/Github/ebnm')
library(pheatmap)
library(ggplot2)
library(symebcovmf)
library(flashier)
```

```{r}
source('code/visualization_functions.R')
```

# Data Generation

```{r}
sim_4pops <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(7 * p, sd = rep(args$branch_sds, each = p)), ncol = 7)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  LL <- matrix(0, nrow = n, ncol = 7)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 1, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 0, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(1, 0, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 1, 0, 0), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 1, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args = list(pop_sizes = rep(40, 4), n_genes = 1000, branch_sds = rep(2,7), indiv_sd = 1, seed = 1)
sim_data <- sim_4pops(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(sim_data$LL, pop_vec)
```

# Initialization

This is the code for the point-Laplace fit plus splitting initialization procedure:
```{r}
laplace_split_initialization <- function(S, Kmax, verbose = 2, backfit_maxiter = 500, backfit_tol = NULL){
  # backfitting is important in the tree setting
  
  # fit point-laplace fit with flash
  flash_laplace_fit <- flash_init(data = S, var_type = 0) |>
    flash_set_verbose(verbose = verbose) |>
    flash_greedy(Kmax = Kmax, ebnm_fn = ebnm::ebnm_point_laplace) |>
    flash_backfit(maxiter = backfit_maxiter, tol = backfit_tol) |>
    flash_nullcheck()
  
  # rescale fit so that L and F are of the same scale
  flash_laplace_fit_scaled <- ldf(flash_laplace_fit, 'i')
  LL <- flash_laplace_fit_scaled$L
  ##FF <- flash_laplace_fit_scaled$F
  
  # split into positive and negative components
  LL <- cbind(pmax(LL, 0), pmax(-LL, 0))
  ##FF <- cbind(pmax(FF, 0), pmax(-FF, 0))
  
  # remove columns of zeros
  idx.nonzero <- apply(LL, 2, function(x){return(sum(x^2))}) > 10^(-10)
  LL <- LL[, idx.nonzero]
  
  # refit weights by least squares
  n <- nrow(S)
  llt_vec <- matrix(rep(0, ncol(LL)*n*n), ncol = ncol(LL))
  for (i in 1:ncol(LL)){
    llt_vec[,i] <- c(LL[,i]%*%t(LL[,i]))
  }

  nnlm_fit <- NNLM::nnlm(llt_vec, as.matrix(c(S), ncol = 1), alpha = c(0,0,0))

  indices_keep <- (nnlm_fit$coefficients > 0)
  LL_scaled <- LL[,indices_keep] %*% diag(sqrt(nnlm_fit$coefficients[indices_keep]))
  
  return(LL_scaled)
}
```

```{r}
LL_split_init <- laplace_split_initialization(sim_data$YYt, Kmax = 7)
```

This is a plot of the initialization:
```{r}
plot_loadings(LL_split_init, rep(c('A','B','C','D'), each = 40))
```

We can see that the initialization contains the components of a tree -- there is an intercept components, two branch components, and four population components.

# symEBcovMF backfit with point-exponential prior

Now, we backfit with symEBcovMF, initialized with the point-Laplace plus splitting estimate.

First, we initialize a symEBcovMF object using $\hat{L}_{split}$.
```{r}
symebcovmf_laplace_split_init_obj <- sym_ebcovmf_init(sim_data$YYt)

split_L_norms <- apply(LL_split_init, 2, function(x){sqrt(sum(x^2))})
idx.keep <- which(split_L_norms > 0)
split_L_normalized <- apply(LL_split_init[,idx.keep], 2, function(x){x/sqrt(sum(x^2))})
init_lambda <- rep(split_L_norms^2, times = 2)[idx.keep]

symebcovmf_laplace_split_init_obj <- sym_ebcovmf_factors_init(sim_data$YYt, symebcovmf_laplace_split_init_obj, split_L_normalized, init_lambda)
```

I refit the lambda values keeping the factors fixed.
```{r}
symebcovmf_laplace_split_obj <- refit_lambda(S = sim_data$YYt, symebcovmf_laplace_split_init_obj, maxiter = 500)
```

```{r, eval = FALSE, include = FALSE}
plot_loadings(symebcovmf_laplace_split_init_obj$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_obj$lambda)), rep(c('A','B','C','D'), each = 40))
```

Now, we run the backfit with point-exponential prior.
```{r, results = 'hide', warning = FALSE}
symebcovmf_laplace_split_init_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_laplace_split_obj, ebnm_fn = ebnm::ebnm_point_exponential, backfit_maxiter = 5000)
```

This is a plot of the estimate from the backfit initialized with the split estimate, $\hat{L}_{split-backfit}$.
```{r}
plot_loadings(symebcovmf_laplace_split_init_backfit$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_backfit$lambda)), rep(c('A','B','C','D'), each = 40))
```

The final estimate looks like a tree. The factors could be more sparse, but overall, the components correspond to the components of a tree.

This is a heatmap of the estimated Gram matrix, $\hat{L}_{split-backfit}\hat{\Lambda}\hat{L}_{split-backfit}'$.
```{r}
gram_est <- tcrossprod(symebcovmf_laplace_split_init_backfit$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_backfit$lambda)))
plot_heatmap(gram_est, brks = seq(0, max(abs(gram_est)), length.out = 50))
```

This is the L2 norm of the difference, excluding diagonal entries.
```{r}
compute_L2_fit <- function(est, dat){
  score <- sum((dat - est)^2) - sum((diag(dat) - diag(est))^2)
  return(score)
}

compute_L2_fit(gram_est, sim_data$YYt)
```

This is a plot of fitted vs. observed values:
```{r}
ggplot(data = NULL, aes(x = as.vector(sim_data$YYt), y = as.vector(gram_est))) + geom_point() + geom_abline(slope = 1, intercept = 0, color = 'red') + xlab('Observed Values') + ylab('Fitted Values')
```

This is a plot of the ELBO:
```{r}
plot(symebcovmf_laplace_split_init_backfit$backfit_iter_elbo_vec)
```

Overall, in this balanced tree setting, symEBcovMF seems to perform well with the point-Laplace plus splitting initialization.
