---
title: "symebcovmf_backfit_debugging_case1"
author: "Annie Xie"
date: "2026-06-27"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, I am working on debugging the backfit function for symebcovmf. While running the DSC, I ran into an issue with the symebcovmf backfit code. This file is for recording the debugging process, identifying the issues, and testing solutions.

# Visualization Functions

```{r}
library(pheatmap)
library(ggplot2)
```

```{r}
plot_heatmap <- function(L, title = "", colors_range = c("gray96", "red"), brks = NULL){
  ### define the color map
  cols <- colorRampPalette(colors_range)(49)
  if (is.null(brks) == TRUE){
    brks <- seq(min(L), max(L), length=50)
  }
  
  plt <- pheatmap(L, show_rownames = FALSE, show_colnames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, color = cols, breaks = brks, main = title)
  return(plt)
}
```

# Setting with Issues
There are a couple of settings where running the symebcovmf backfit yields an error about NaN standard errors in the EBNM problem. I will try to recreate that issue here.

I load in one of the sparse, overlapping groups datasets.
```{r}
group_overlap_5 <- readRDS('data/group_overlap_5.rds')
```

```{r}
library(symebcovmf)
```

First, we run greedy symebcovmf with `Kmax = 20` (double the true number of factors) and the generalized binary prior.
```{r}
symebcovmf_greedy_fit <- sym_ebcovmf_fit(S = group_overlap_5$data$YYt,
                                         ebnm_fn = ebnm::ebnm_generalized_binary,
                                         Kmax = 20,
                                         maxiter = 500,
                                         rank_one_tol = 10^(-8),
                                         tol = 10^(-8),
                                         refit_lam = TRUE)
```

This is the backfit function:
```{r}
sym_ebcovmf_backfit <- function(S, sym_ebcovmf_obj, ebnm_fn, backfit_maxiter = 100, backfit_tol = 10^(-8), optim_maxiter= 500, optim_tol = 10^(-8)){
  K <- length(sym_ebcovmf_obj$lambda)
  kset <- c(1:K)
  iter <- 1
  obj_diff <- Inf
  sym_ebcovmf_obj$backfit_vec_elbo_full <- NULL
  sym_ebcovmf_obj$backfit_iter_elbo_vec <- NULL
  
  # refit lambda
  # sym_ebcovmf_obj <- refit_lambda(S, sym_ebcovmf_obj, maxiter = 25)
  
  while((iter <= backfit_maxiter) && (obj_diff > backfit_tol)){
    # print(iter)
    obj_old <- sym_ebcovmf_obj$elbo
    # loop through each factor
    for (k in kset){
      # print(k)
      print(paste('Updating factor', k))
      # compute residual matrix
      R <- S - tcrossprod(sym_ebcovmf_obj$L_pm[,-k, drop = FALSE] %*% diag(sqrt(sym_ebcovmf_obj$lambda[-k]), ncol = (K-1)))
      R2k <- compute_R2(S, sym_ebcovmf_obj$L_pm[,-k, drop = FALSE], sym_ebcovmf_obj$lambda[-k], (K-1)) #this is right but I have one instance where the values don't match what I expect
      
      # optimize factor
      factor_proposed <- optimize_factor(R, ebnm_fn, optim_maxiter, optim_tol, sym_ebcovmf_obj$L_pm[,k], sym_ebcovmf_obj$lambda[k], sym_ebcovmf_obj$fitted_gs[[k]], R2k, sym_ebcovmf_obj$n, sym_ebcovmf_obj$KL[-k])
      
      # update object
      # check if update leads to increase in objective function
      if ((factor_proposed$curr_elbo > sym_ebcovmf_obj$elbo) | (iter == 1)){
        sym_ebcovmf_obj$L_pm[,k] <- factor_proposed$v
        sym_ebcovmf_obj$KL[k] <- factor_proposed$rank_one_KL
        sym_ebcovmf_obj$lambda[k] <- factor_proposed$lambda_k
        sym_ebcovmf_obj$resid_s2 <- factor_proposed$resid_s2
        sym_ebcovmf_obj$fitted_gs[[k]] <- factor_proposed$fitted_g_k
        sym_ebcovmf_obj$elbo <- factor_proposed$curr_elbo
        sym_ebcovmf_obj$backfit_vec_elbo_full <- c(sym_ebcovmf_obj$backfit_vec_elbo_full, factor_proposed$vec_elbo_full)
      } else {
        obj_diff <- sym_ebcovmf_obj$elbo - factor_proposed$curr_elbo
        print(paste('update to factor', k, 'decreased the elbo by', abs(obj_diff)))
      }
      
      #print(sym_ebcovmf_obj$elbo)
      # sym_ebcovmf_obj <- refit_lambda(S, sym_ebcovmf_obj, maxiter = 1, remove_null = FALSE) # add refitting step?
      # print(sym_ebcovmf_obj$lambda)
      #print(sym_ebcovmf_obj$elbo)
    }
    kset <- which(sym_ebcovmf_obj$lambda != 0)
    sym_ebcovmf_obj$backfit_iter_elbo_vec <- c(sym_ebcovmf_obj$backfit_iter_elbo_vec, sym_ebcovmf_obj$elbo)
    
    iter <- iter + 1
    obj_diff <- abs(sym_ebcovmf_obj$elbo - obj_old)
  }
  # nullcheck
  sym_ebcovmf_obj <- nullcheck_factors(S, sym_ebcovmf_obj)
  return(sym_ebcovmf_obj)
}
```

This is the code to run the backfit:
```{r, eval = FALSE}
symebcovmf_backfit_fit <- sym_ebcovmf_backfit(S = group_overlap_5$data$YYt, 
                                              sym_ebcovmf_obj = symebcovmf_greedy_fit, 
                                              ebnm_fn = ebnm::ebnm_generalized_binary, 
                                              backfit_maxiter = 500, 
                                              backfit_tol = 10^(-8), 
                                              optim_maxiter = 500, 
                                              optim_tol = 10^(-8))
```

When I run the backfit, I get an error about missing standard errors in the EBNM solver.

(From my handwritten notes): From my previous debugging, I found that these errors were due to $v^{T}Rv$ being negative. If $R$ were positive semi-definite, then we wouldn't have to worry about this issue. But since $R$ is a residual matrix, we have no guarantees on the matrix being positive semi-definite. 

An arguably more principled way to fix this is to maybe add a constraint to the optimization, e.g. solve for $v$ such that $R - \lambda v v^{T}$ is positive semi-definite. A more ad-hoc work-around is to, e.g., refit the weights after each factor update in the backfit. If a weight goes negative, then it will just get set to zero and that factor will be removed, circumventing the issue of non-defined standard errors.

# Fixing the Backfit
We try running the proposed fix in this section.
```{r}
source('code/symebcovmf_debugging_script.R')
```

```{r}
symebcovmf_backfit_fit <- sym_ebcovmf_backfit_alt(S = group_overlap_5$data$YYt, 
                                              sym_ebcovmf_obj = symebcovmf_greedy_fit, 
                                              ebnm_fn = ebnm::ebnm_generalized_binary, 
                                              backfit_maxiter = 1000, 
                                              backfit_tol = 10^(-8), 
                                              optim_maxiter = 500, 
                                              optim_tol = 10^(-8))
```

# New Backfit Results
These are the results. This is the number of factors in the loadings estimate.
```{r}
ncol(symebcovmf_backfit_fit$L_pm)
```

This is a heatmap of the loadings estimate:
```{r}
plot_heatmap(symebcovmf_backfit_fit$L_pm %*% diag(sqrt(symebcovmf_backfit_fit$lambda)))
```

The modified backfit runs in this setting.