---
title: "symebcovmf_point_exp_backfit_tree"
author: "Annie Xie"
date: "2025-06-17"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, we explore symEBcovMF (with backfit) with the point-exponential prior in the tree setting.

## Motivation
I am interested in comparing how greedy symEBcovMF + backfit with the point-exponential prior compares with (backfit) symEBcovMF with the point-Laplace plus splitting initialization. The point-Laplace plus splitting initialization procedure is used in GBCD, and we've found it to work well empirically. However, it is a complicated procedure. Therefore, I want to see if the simpler procedure of greedy symEBcovMF + backfit with the point-exponential prior also works.

# Packages and Functions

```{r}
library(ebnm)
library(pheatmap)
library(ggplot2)
library(symebcovmf)
```

```{r}
source('code/visualization_functions.R')
```

```{r}
compute_L2_fit <- function(est, dat){
  score <- sum((dat - est)^2) - sum((diag(dat) - diag(est))^2)
  return(score)
}
```

```{r}
compute_crossprod_similarity <- function(est, truth){
  K_est <- ncol(est)
  K_truth <- ncol(truth)
  n <- nrow(est)
  
  #if estimates don't have same number of columns, try padding the estimate with zeros and make cosine similarity zero
  if (K_est < K_truth){
    est <- cbind(est, matrix(rep(0, n*(K_truth-K_est)), nrow = n))
  }
  
  if (K_est > K_truth){
    truth <- cbind(truth, matrix(rep(0, n*(K_est - K_truth)), nrow = n))
  }
  
  #normalize est and truth
  norms_est <- apply(est, 2, function(x){sqrt(sum(x^2))})
  norms_est[norms_est == 0] <- Inf
  
  norms_truth <- apply(truth, 2, function(x){sqrt(sum(x^2))})
  norms_truth[norms_truth == 0] <- Inf
  
  est_normalized <- t(t(est)/norms_est)
  truth_normalized <- t(t(truth)/norms_truth)
  
  #compute matrix of cosine similarities
  cosine_sim_matrix <- abs(crossprod(est_normalized, truth_normalized))
  
  assignment_problem <- lpSolve::lp.assign(t(cosine_sim_matrix), direction = "max")
  return((1/K_truth)*assignment_problem$objval)
}
```

# Data Generation
To test this procedure, I will apply it to the tree-structured dataset.

```{r}
sim_4pops <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(7 * p, sd = rep(args$branch_sds, each = p)), ncol = 7)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  LL <- matrix(0, nrow = n, ncol = 7)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 1, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 0, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(1, 0, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 1, 0, 0), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 1, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args = list(pop_sizes = rep(40, 4), n_genes = 1000, branch_sds = rep(2,7), indiv_sd = 1, seed = 1)
sim_data <- sim_4pops(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(sim_data$LL, pop_vec)
```

This is a plot of the eigenvalues of the Gram matrix:
```{r}
S_eigen <- eigen(sim_data$YYt)
plot(S_eigen$values) + abline(a = 0, b = 0, col = 'red')
```

This is the minimum eigenvalue:
```{r}
min(S_eigen$values)
```

# symEBcovMF with point-exponential

First, we start with running greedy symEBcovMF with the point-exponential prior.
```{r}
symebcovmf_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm::ebnm_point_exponential, K = 7, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}_{pt-exp}$, the estimate from symEBcovMF:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_fit$L_pm %*% diag(sqrt(symebcovmf_fit$lambda)), bal_pops)
```

This is the objective function value attained:
```{r}
symebcovmf_fit$elbo
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit$L_pm, sim_data$LL)
```

## Observations
The estimate from greedy symEBcovMF with point-exponential prior generally looks like a tree -- there is one intercept factor, two factors corresponding to branch effects, and four factors corresponding primarily to population effects. Some of the population effect factors have very small non-zero loadings for members in other populations. But overall, the estimate does a good job at capturing the main tree components.

## Additional Backfit

Now, we run additional backfit. This is the code for the backfit:
```{r, results = 'hide'}
symebcovmf_fit_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_fit, ebnm_fn = ebnm_point_exponential, backfit_maxiter = 500)
```

This is a scatter plot of $\hat{L}_{pt-exp-backfit}$, the estimate from symEBcovMF with backfit:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_fit_backfit$L_pm %*% diag(sqrt(symebcovmf_fit_backfit$lambda)), bal_pops)
```

This is the objective function value attained:
```{r}
symebcovmf_fit_backfit$elbo
```

This is a plot of the progression of the ELBO:
```{r}
plot(symebcovmf_fit_backfit$backfit_iter_elbo_vec, ylab = 'ELBO')
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit_backfit$L_pm, sim_data$LL)
```

Interestingly, the backfit made the factors less sparse. I would expect the backfit to find sparser representations since I expect the sparse representation to have a higher objective function value. I also saw these small non-zero loadings in the estimates from GBCD and symEBcovMF initialized with the point-Laplace fit plus splitting. Perhaps they are a result of model misspecification with regards to the noise?

Note: I tried running the backfit for more iterations, e.g. 10,000 iterations, and found that the estimate looked similar. A small number of samples had changed loading values in each factor, but the main components remained intact.

# symEBcovMF backfit initialized with true loadings value
To test if symEBcovMF prefers the estimate with small non-zero loadings, I try running the backfit from the true loadings matrix. I suspect we will see an estimate similar to that of the previous section since the greedy symEBcovMF estimate was pretty similar to the true loadings matrix.

First, we initialize a symEBcovMF object using the true $L$.
```{r}
symebcovmf_true_init_obj <- sym_ebcovmf_init(sim_data$YYt)

true_L_normalized <- apply(sim_data$LL, 2, function(x){x/sqrt(sum(x^2))})
symebcovmf_true_init_obj$L_pm <- true_L_normalized

symebcovmf_true_init_obj$lambda <- rep(1, ncol(sim_data$LL))

symebcovmf_true_init_obj$resid_s2 <- estimate_resid_s2(S = sim_data$YYt,
  L = symebcovmf_true_init_obj$L_pm, 
  lambda = symebcovmf_true_init_obj$lambda, 
  n = nrow(sim_data$Y), 
  K = length(symebcovmf_true_init_obj$lambda))

symebcovmf_true_init_obj$elbo <- compute_elbo(S = sim_data$YYt, 
  L = symebcovmf_true_init_obj$L_pm, 
  lambda = symebcovmf_true_init_obj$lambda, 
  resid_s2 = symebcovmf_true_init_obj$resid_s2, 
  n = nrow(sim_data$Y), 
  K = length(symebcovmf_true_init_obj$lambda), 
  KL = rep(0, length(symebcovmf_true_init_obj$lambda)))

symebcovmf_true_init_obj$fitted_gs <- rep(list(NULL), length(symebcovmf_true_init_obj$lambda))
```

I refit the lambda values keeping the factors fixed.
```{r}
symebcovmf_true_init_obj <- refit_lambda(S = sim_data$YYt, symebcovmf_true_init_obj, maxiter = 500)
```

Now, we run the backfit with point-exponential prior.
```{r, results = 'hide'}
symebcovmf_true_init_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_true_init_obj, ebnm_fn = ebnm::ebnm_point_exponential, backfit_maxiter = 500)
```

This is a plot of the loadings estimate, $\hat{L}_{backfit-true-init}$.
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_true_init_backfit$L_pm %*% diag(sqrt(symebcovmf_true_init_backfit$lambda)), bal_pops)
```

This is the objective function value:
```{r}
symebcovmf_true_init_backfit$elbo
```

This is a plot of the progression of the ELBO:
```{r}
plot(symebcovmf_true_init_backfit$backfit_iter_elbo_vec, ylab = 'ELBO')
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_true_init_backfit$L_pm, sim_data$LL)
```

## Observations
The factors in this estimate also have small non-zero loadings. These results suggest that the less-sparse representation has a higher objective function value.
