---
title: "symebcovmf_tree_exploration"
author: "Annie Xie"
date: "2025-04-15"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this example, we further explore symEBcovMF on tree data.

# Packages and Functions

```{r}
library(ebnm)
library(pheatmap)
library(ggplot2)
```

```{r}
plot_heatmap <- function(L, title = "", colors_range = c("gray96", "red"), brks = NULL){
  ### define the color map
  cols <- colorRampPalette(colors_range)(49)
  if (is.null(brks) == TRUE){
    brks <- seq(min(L), max(L), length=50)
  }
  
  plt <- pheatmap(L, show_rownames = FALSE, show_colnames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, color = cols, breaks = brks, main = title)
  return(plt)
}

plot_loadings <- function(L_est, Pop, legendYN = TRUE, scales_option = "fixed"){
  # options for scales_option are "fixed" and "free_y"
  n <- nrow(L_est)
  k <- ncol(L_est)
  Idx <- rep(c(1:n), k)
  Loading <- c(L_est)
  Factor <- paste0('k=',c(sapply(c(1:k), function(x, n){rep(x, n)}, n = n)))
  tib <- data.frame(Idx, Loading, Factor, Pop)
  plt <- ggplot(tib, aes(x = Idx, y = Loading, col = Pop)) +
    geom_point(show.legend = legendYN) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    facet_grid(cols = vars(Factor), scales = scales_option) +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          panel.spacing = unit(1, "lines"))
  plot(plt)
}
```

```{r}
source('code/symebcovmf_functions.R')
```

# Data Generation

```{r}
sim_4pops <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(7 * p, sd = rep(args$branch_sds, each = p)), ncol = 7)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  LL <- matrix(0, nrow = n, ncol = 7)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 1, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 0, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(1, 0, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 1, 0, 0), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 1, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args = list(pop_sizes = rep(40, 4), n_genes = 1000, branch_sds = rep(2,7), indiv_sd = 1, seed = 1)
sim_data <- sim_4pops(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(sim_data$LL, pop_vec)
```

This is a plot of the eigenvalues of the Gram matrix:
```{r}
S_eigen <- eigen(sim_data$YYt)
plot(S_eigen$values) + abline(a = 0, b = 0, col = 'red')
```

This is the minimum eigenvalue:
```{r}
min(S_eigen$values)
```

# symEBcovMF Set Up

I am particularly interested in when the fourth factor is fit. So I will start by fitting three factors.

```{r}
symebcovmf_exp_init_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm::ebnm_point_exponential, K = 3, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

```{r}
symebcovmf_gb_init_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm::ebnm_generalized_binary, K = 3, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}_{exp}$, the estimate from symEBcovMF with point-exp prior:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_exp_init_fit$L_pm %*% diag(sqrt(symebcovmf_exp_init_fit$lambda)), bal_pops)
```

This is a scatter plot of $\hat{L}_{gb}$, the estimate from symEBcovMF with gb prior:
```{r}
plot_loadings(symebcovmf_gb_init_fit$L_pm %*% diag(sqrt(symebcovmf_gb_init_fit$lambda)), bal_pops)
```

This is a plot of the eigenvalues of the residual matrix for point-exp prior:
```{r}
resid_exp_eigen <- eigen(sim_data$YYt - tcrossprod(symebcovmf_exp_init_fit$L_pm %*% diag(sqrt(symebcovmf_exp_init_fit$lambda), ncol = 1)))
plot(resid_exp_eigen$values) + abline(h=0, col = 'red')
```

```{r}
min(resid_exp_eigen$values)
```

This is a plot of the eigenvalues of the residual matrix for gb prior:
```{r}
resid_gb_eigen <- eigen(sim_data$YYt - tcrossprod(symebcovmf_gb_init_fit$L_pm %*% diag(sqrt(symebcovmf_gb_init_fit$lambda), ncol = 1)))
plot(resid_gb_eigen$values) + abline(h=0, col = 'red')
```

```{r}
min(resid_gb_eigen$values)
```

# Fitting the fourth factor

## Point-exponential prior
```{r}
symebcovmf_exp_fit <- sym_ebcovmf_r1_fit(sim_data$YYt, symebcovmf_exp_init_fit, ebnm_fn = ebnm::ebnm_point_exponential, maxiter = 500, tol = 10^(-8))
symebcovmf_exp_fit <- refit_lambda(sim_data$YYt, symebcovmf_exp_fit)
```

### Visualization of Estimate
```{r}
plot(symebcovmf_exp_fit$L_pm[,4])
```

### Visualization of ELBO

This is a plot of the progression of the ELBO:
```{r}
idx <- which(symebcovmf_exp_fit$vec_elbo_full == 4)
plot(symebcovmf_exp_fit$vec_elbo_full[-c(1:idx)], type = 'b', xlab = 'Iter', ylab = 'ELBO') 
```

### Progression of Estimate
```{r}
R <- sim_data$YYt - tcrossprod(symebcovmf_exp_init_fit$L_pm %*% diag(sqrt(symebcovmf_exp_init_fit$lambda), ncol = 3))
estimates_exp_list <- list(sym_ebcovmf_r1_init(R)$v)
for (i in 1:13){
  estimates_exp_list[[(i+1)]] <- sym_ebcovmf_r1_fit(sim_data$YYt, symebcovmf_exp_init_fit, ebnm_fn = ebnm::ebnm_point_exponential, maxiter = i, tol = 10^(-8))$L_pm[,4]
}
```

```{r}
par(mfrow = c(7,2), mar = c(2, 2, 1, 1) + 0.1)
for (i in 1:14){
  plot(estimates_exp_list[[i]], main = paste('Iter', (i-1)), ylab = 'L')
}
par(mfrow = c(1,1))
```

## Generalized binary prior
```{r}
symebcovmf_gb_fit <- sym_ebcovmf_r1_fit(sim_data$YYt, symebcovmf_gb_init_fit, ebnm_fn = ebnm::ebnm_generalized_binary, maxiter = 500, tol = 10^(-8))
symebcovmf_gb_fit <- refit_lambda(sim_data$YYt, symebcovmf_gb_fit)
```

### Visualization of Estimate
```{r}
plot(symebcovmf_gb_fit$L_pm[,4])
```

### Visualization of ELBO

This is a plot of the progression of the ELBO:
```{r}
idx <- which(symebcovmf_gb_fit$vec_elbo_full == 4)
plot(symebcovmf_gb_fit$vec_elbo_full[-c(1:idx)], type = 'b', xlab = 'Iter', ylab = 'ELBO') 
```

### Progression of Estimate
```{r}
R <- sim_data$YYt - tcrossprod(symebcovmf_gb_init_fit$L_pm %*% diag(sqrt(symebcovmf_gb_init_fit$lambda), ncol = 3))
estimates_gb_list <- list(sym_ebcovmf_r1_init(R)$v)
for (i in 1:11){
  estimates_gb_list[[(i+1)]] <- sym_ebcovmf_r1_fit(sim_data$YYt, symebcovmf_gb_init_fit, ebnm_fn = ebnm::ebnm_generalized_binary, maxiter = i, tol = 10^(-8))$L_pm[,4]
}
```

```{r}
par(mfrow = c(6,2), mar = c(2, 2, 1, 1) + 0.1)
max_y <- max(sapply(estimates_gb_list, max))
for (i in 1:12){
  plot(estimates_gb_list[[i]], main = paste('Iter', (i-1)), ylab = 'L', ylim = c(0, max_y))
}
par(mfrow = c(1,1))
```

## Observations
We can see that point-exponential prior is able to find a sparser solution in this setting. On the other hand, the generalized binary prior found a less sparse, binary solution. Something to try in the future: if I initialize the generalized-binary prior fit with the correct factor, will it recover something close to the correct factor? General question: how sensitive is the generalized binary prior to initialization? My guess is it would be pretty sensitive.