---
title: "symebcovmf_laplace_split_init_tree_exploration"
author: "Annie Xie"
date: "2025-06-19"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, I further explore the point-Laplace plus splitting initialization strategy for symEBcovMF in the tree setting.

# Packages and Functions

```{r}
library(ebnm)
library(pheatmap)
library(ggplot2)
library(symebcovmf)
```

```{r}
source('code/visualization_functions.R')
```

```{r}
compute_crossprod_similarity <- function(est, truth){
  K_est <- ncol(est)
  K_truth <- ncol(truth)
  n <- nrow(est)
  
  #if estimates don't have same number of columns, try padding the estimate with zeros and make cosine similarity zero
  if (K_est < K_truth){
    est <- cbind(est, matrix(rep(0, n*(K_truth-K_est)), nrow = n))
  }
  
  if (K_est > K_truth){
    truth <- cbind(truth, matrix(rep(0, n*(K_est - K_truth)), nrow = n))
  }
  
  #normalize est and truth
  norms_est <- apply(est, 2, function(x){sqrt(sum(x^2))})
  norms_est[norms_est == 0] <- Inf
  
  norms_truth <- apply(truth, 2, function(x){sqrt(sum(x^2))})
  norms_truth[norms_truth == 0] <- Inf
  
  est_normalized <- t(t(est)/norms_est)
  truth_normalized <- t(t(truth)/norms_truth)
  
  #compute matrix of cosine similarities
  cosine_sim_matrix <- abs(crossprod(est_normalized, truth_normalized))
  
  assignment_problem <- lpSolve::lp.assign(t(cosine_sim_matrix), direction = "max")
  return((1/K_truth)*assignment_problem$objval)
}
```

# Data Generation
To test this procedure, I will apply it to the tree-structured dataset.

```{r}
sim_4pops <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(7 * p, sd = rep(args$branch_sds, each = p)), ncol = 7)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  LL <- matrix(0, nrow = n, ncol = 7)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 1, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 0, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(1, 0, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 1, 0, 0), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 1, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args = list(pop_sizes = rep(40, 4), n_genes = 1000, branch_sds = rep(2,7), indiv_sd = 1, seed = 1)
sim_data <- sim_4pops(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(sim_data$LL, pop_vec)
```

This is a plot of the eigenvalues of the Gram matrix:
```{r}
S_eigen <- eigen(sim_data$YYt)
plot(S_eigen$values) + abline(a = 0, b = 0, col = 'red')
```

This is the minimum eigenvalue:
```{r}
min(S_eigen$values)
```

# Split point-Laplace Fit

I will load in previously saved results for the point-Laplace fit:
```{r}
symebcovmf_laplace_fit <- readRDS('data/symebcovmf_laplace_tree_backfit_20000.rds')
```

Now, we split the estimate into its positive and negative parts.
```{r}
LL_split_init <- cbind(pmax(symebcovmf_laplace_fit$L_pm, 0), pmax(-symebcovmf_laplace_fit$L_pm, 0))
```

This is a plot of the estimate with its positive and negative parts split into different factors, $\hat{L}_{split}$.
```{r}
plot_loadings(LL_split_init, pop_vec)
```

This is a heatmap of the estimate with its positive and negative parts split into different factors, $\hat{L}_{split}$.
```{r}
plot_heatmap(LL_split_init)
```

# Backfit with Split Initialization
Now, we backfit from the split initialization (this is what GBCD does). The backfit is run with the point-exponential prior.

First, we initialize a symEBcovMF object using $\hat{L}_{split}$.
```{r}
symebcovmf_laplace_split_init_obj <- sym_ebcovmf_init(sim_data$YYt)

split_L_norms <- apply(LL_split_init, 2, function(x){sqrt(sum(x^2))})
idx.keep <- which(split_L_norms > 0)
split_L_normalized <- apply(LL_split_init[,idx.keep], 2, function(x){x/sqrt(sum(x^2))})
symebcovmf_laplace_split_init_obj$L_pm <- split_L_normalized

symebcovmf_laplace_split_init_obj$lambda <- rep(symebcovmf_laplace_fit$lambda, times = 2)[idx.keep]

symebcovmf_laplace_split_init_obj$resid_s2 <- estimate_resid_s2(S = sim_data$YYt, 
  L = symebcovmf_laplace_split_init_obj$L_pm, 
  lambda = symebcovmf_laplace_split_init_obj$lambda, 
  n = nrow(sim_data$Y), 
  K = length(symebcovmf_laplace_split_init_obj$lambda))

symebcovmf_laplace_split_init_obj$elbo <- compute_elbo(S = sim_data$YYt, 
  L = symebcovmf_laplace_split_init_obj$L_pm, 
  lambda = symebcovmf_laplace_split_init_obj$lambda, 
  resid_s2 = symebcovmf_laplace_split_init_obj$resid_s2, 
  n = nrow(sim_data$Y), 
  K = length(symebcovmf_laplace_split_init_obj$lambda), 
  KL = rep(0, length(symebcovmf_laplace_split_init_obj$lambda)))

symebcovmf_laplace_split_init_obj$fitted_gs <- rep(list(NULL), length(symebcovmf_laplace_split_init_obj$lambda))
```

I refit the lambda values keeping the factors fixed.
```{r}
symebcovmf_laplace_split_obj <- refit_lambda(S = sim_data$YYt, symebcovmf_laplace_split_init_obj, maxiter = 500)
```

This is a plot of the initial loadings estimate:
```{r}
plot_loadings(symebcovmf_laplace_split_init_obj$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_obj$lambda)), pop_vec)
```

This is a plot of the initial estimate of the scaled Gram matrix:
```{r}
plot_heatmap(tcrossprod(symebcovmf_laplace_split_init_obj$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_obj$lambda))), brks = seq(0, max(abs(tcrossprod(symebcovmf_laplace_split_init_obj$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_obj$lambda))))), length.out = 50))
```

Now, we run the backfit with point-exponential prior.
```{r, results = 'hide'}
symebcovmf_laplace_split_init_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_laplace_split_init_obj, ebnm_fn = ebnm::ebnm_point_exponential, backfit_maxiter = 500)
```

This is a plot of the estimate from the backfit initialized with the split estimate, $\hat{L}_{split-backfit}$.
```{r}
plot_loadings(symebcovmf_laplace_split_init_backfit$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_backfit$lambda)), pop_vec)
```

This is a heatmap of the estimate from the backfit initialized with the split estimate, $\hat{L}_{split-backfit}$.
```{r}
plot_heatmap(symebcovmf_laplace_split_init_backfit$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_backfit$lambda)))
```

This is the crossproduct similarity of the loadings estimate from the backfit initialized with the split estimate, $\hat{L}_{split-backfit}$.
```{r}
compute_crossprod_similarity(symebcovmf_laplace_split_init_backfit$L_pm, sim_data$LL)
```

This is a heatmap of the estimated Gram matrix, $\hat{L}_{split-backfit}\hat{\Lambda}\hat{L}_{split-backfit}'$.
```{r}
gram_est <- tcrossprod(symebcovmf_laplace_split_init_backfit$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_backfit$lambda)))
plot_heatmap(gram_est, brks = seq(0, max(abs(gram_est)), length.out = 50))
```

This is the L2 norm of the difference, excluding diagonal entries.
```{r}
compute_L2_fit <- function(est, dat){
  score <- sum((dat - est)^2) - sum((diag(dat) - diag(est))^2)
  return(score)
}

compute_L2_fit(gram_est, sim_data$YYt)
```

This is a plot of fitted vs. observed values:
```{r}
ggplot(data = NULL, aes(x = as.vector(sim_data$YYt), y = as.vector(gram_est))) + geom_point() + geom_abline(slope = 1, intercept = 0, color = 'red') + xlab('Observed Values') + ylab('Fitted Values')
```

This is a plot of the progression of the ELBO:
```{r}
plot(symebcovmf_laplace_split_init_backfit$backfit_iter_elbo_vec, ylab = 'ELBO')
```

## Observations
The estimate from the backfit still captures the tree components and yields a very high crossproduct similarity value. One observation is some of the factors are not fully sparse. For example, in factor 3, which primarily captures the the population effect for population 3, we see that members of populations 1 and 2 now have small non-zero loading. Based on my intuition, I would expect the fully sparse representation to have a higher objective function value, so I'm not exactly sure why these factors have these small loading values in other populations. I imagine these small loading values could be shrunk to zero if you used a binary (or a generalized binary prior that leans toward being binary) to improve the fit.

# GBCD
For comparison, I also try applying GBCD with point-exponential prior.

```{r}
library(gbcd)
```

```{r}
gbcd_fit <- fit_gbcd(sim_data$Y, Kmax = 7, ebnm::ebnm_point_exponential)
```

This is a plot of the loadings estimate from GBCD, $\hat{L}_{gbcd}$.
```{r}
plot_loadings(gbcd_fit$L, pop_vec)
```

This is a heatmap of the loadings estimate from GBCD, $\hat{L}_{gbcd}$.
```{r}
plot_heatmap(gbcd_fit$L)
```

This is the crossproduct similarity of the loadings estimate from GBCD, $\hat{L}_{gbcd}$.
```{r}
compute_crossprod_similarity(gbcd_fit$L, sim_data$LL)
```

## Observations
The estimate from GBCD also captures all the tree components and yields a high crossproduct similarity value. The crossproduct similarity is just slightly lower than that of symEBcovMF backfit initialized with the point-Laplace fit plus splitting. We also find that some of the GBCD factors have small non-zero loadings. For example, factor 5, which primarily captures population 2, has small non-zero loadings for members of populations 3 and 4. Given that the GBCD estimate also has this characteristic, I am inclined to think non-zero loadings are a result of model misspecification.
