---
title: "symebcovmf_laplace_split_init_tree"
author: "Annie Xie"
date: "2025-05-15"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, we explore symEBcovMF with the GBCD initialization procedure -- fitting with point-Laplace prior and then splitting the matrices into the negative and non-negative parts -- in the tree setting.

## Motivation
We are interested in this procedure because it is the initialization procedure for GBCD, and it seems to work well empirically. We are interested in whether this procedure also works well for symEBcovMF. There isn't much justification for this procedure beyond Jason's work with divergence factorizations in the tree setting.

# Packages and Functions

```{r}
library(ebnm)
library(pheatmap)
library(ggplot2)
library(symebcovmf)
```

```{r}
source('code/visualization_functions.R')
```

# Data Generation
To test this procedure, I will apply it to the tree-structured dataset.

```{r}
sim_4pops <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(7 * p, sd = rep(args$branch_sds, each = p)), ncol = 7)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  LL <- matrix(0, nrow = n, ncol = 7)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 1, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 0, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(1, 0, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 1, 0, 0), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 1, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args = list(pop_sizes = rep(40, 4), n_genes = 1000, branch_sds = rep(2,7), indiv_sd = 1, seed = 1)
sim_data <- sim_4pops(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(sim_data$LL, pop_vec)
```

This is a plot of the eigenvalues of the Gram matrix:
```{r}
S_eigen <- eigen(sim_data$YYt)
plot(S_eigen$values) + abline(a = 0, b = 0, col = 'red')
```

This is the minimum eigenvalue:
```{r}
min(S_eigen$values)
```

# symEBcovMF with point-Laplace

First, we run greedy symEBcovMF with the point-Laplace prior.
```{r}
symebcovmf_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm::ebnm_point_laplace, K = 7, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), sign_constraint = NULL, refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}_{pt-laplace}$, the estimate from symEBcovMF:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_fit$L_pm %*% diag(sqrt(symebcovmf_fit$lambda)), bal_pops)
```

This is the objective function value attained:
```{r}
symebcovmf_fit$elbo
```

## Additional Backfit
GBCD does additional backfitting on the point-Laplace fit, so I want to test how additional backfitting performs.

```{r, results = 'hide'}
symebcovmf_fit_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_fit, ebnm_fn = ebnm_point_laplace, backfit_maxiter = 500)
```

This is a scatter plot of $\hat{L}_{pt-laplace-backfit}$, the estimate from symEBcovMF with backfit:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_fit_backfit$L_pm %*% diag(sqrt(symebcovmf_fit_backfit$lambda)), bal_pops)
```

This is the objective function value attained:
```{r}
symebcovmf_fit_backfit$elbo
```

The estimate from the backfit attains a higher objective function, but is not characteristically very different from the greedy procedure estimate.

# Split
Now, we split the estimate into its positive and negative parts.

```{r}
LL_split_init <- cbind(pmax(symebcovmf_fit$L_pm, 0), pmax(-symebcovmf_fit$L_pm, 0))
```

This is a heatmap of the estimate with its positive and negative parts split into different factors, $\hat{L}_{split}$.
```{r}
plot_heatmap(LL_split_init)
```

# Backfit with Split Initialization
Now, we backfit from the split initialization (this is what GBCD does). The backfit is run with the point-exponential prior.

First, we initialize a symEBcovMF object using $\hat{L}_{split}$.
```{r}
symebcovmf_laplace_split_init_obj <- sym_ebcovmf_init(sim_data$YYt)

split_L_norms <- apply(LL_split_init, 2, function(x){sqrt(sum(x^2))})
idx.keep <- which(split_L_norms > 0)
split_L_normalized <- apply(LL_split_init[,idx.keep], 2, function(x){x/sqrt(sum(x^2))})
symebcovmf_laplace_split_init_obj$L_pm <- split_L_normalized

symebcovmf_laplace_split_init_obj$lambda <- rep(symebcovmf_fit$lambda, times = 2)[idx.keep]

symebcovmf_laplace_split_init_obj$resid_s2 <- estimate_resid_s2(S = sim_data$YYt, 
  L = symebcovmf_laplace_split_init_obj$L_pm, 
  lambda = symebcovmf_laplace_split_init_obj$lambda, 
  n = nrow(sim_data$Y), 
  K = length(symebcovmf_laplace_split_init_obj$lambda))

symebcovmf_laplace_split_init_obj$elbo <- compute_elbo(S = sim_data$YYt, 
  L = symebcovmf_laplace_split_init_obj$L_pm, 
  lambda = symebcovmf_laplace_split_init_obj$lambda, 
  resid_s2 = symebcovmf_laplace_split_init_obj$resid_s2, 
  n = nrow(sim_data$Y), 
  K = length(symebcovmf_laplace_split_init_obj$lambda), 
  KL = rep(0, length(symebcovmf_laplace_split_init_obj$lambda)))

symebcovmf_laplace_split_init_obj$fitted_gs <- rep(list(NULL), length(symebcovmf_laplace_split_init_obj$lambda))
```

I refit the lambda values keeping the factors fixed.
```{r}
symebcovmf_laplace_split_obj <- refit_lambda(S = sim_data$YYt, symebcovmf_laplace_split_init_obj, maxiter = 500)
```

```{r, eval = FALSE, include = FALSE}
plot_loadings(symebcovmf_laplace_split_init_obj$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_obj$lambda)), bal_pops)
```

```{r, eval = FALSE, include = FALSE}
plot_heatmap(tcrossprod(symebcovmf_laplace_split_init_obj$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_obj$lambda))), brks = seq(0, max(abs(tcrossprod(symebcovmf_laplace_split_init_obj$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_obj$lambda))))), length.out = 50))
```

Now, we run the backfit with point-exponential prior.
```{r, results = 'hide'}
symebcovmf_laplace_split_init_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_laplace_split_init_obj, ebnm_fn = ebnm::ebnm_point_exponential, backfit_maxiter = 500)
```

This is a plot of the estimate from the backfit initialized with the split estimate, $\hat{L}_{split-backfit}$.
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_laplace_split_init_backfit$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_backfit$lambda)), bal_pops)
```

This is a heatmap of the estimated Gram matrix, $\hat{L}_{split-backfit}\hat{\Lambda}\hat{L}_{split-backfit}'$.
```{r}
gram_est <- tcrossprod(symebcovmf_laplace_split_init_backfit$L_pm %*% diag(sqrt(symebcovmf_laplace_split_init_backfit$lambda)))
plot_heatmap(gram_est, brks = seq(0, max(abs(gram_est)), length.out = 50))
```

This is the L2 norm of the difference, excluding diagonal entries.
```{r}
compute_L2_fit <- function(est, dat){
  score <- sum((dat - est)^2) - sum((diag(dat) - diag(est))^2)
  return(score)
}

compute_L2_fit(gram_est, sim_data$YYt)
```

This is a plot of fitted vs. observed values:
```{r}
ggplot(data = NULL, aes(x = as.vector(sim_data$YYt), y = as.vector(gram_est))) + geom_point() + geom_abline(slope = 1, intercept = 0, color = 'red') + xlab('Observed Values') + ylab('Fitted Values')
```

# Observations
We see that this procedure does not yield a tree-structured estimate. Instead of four single population effect factors, we have four factors each containing two population effects. The estimated Gram matrix looks close to the observed Gram matrix, suggesting an identifiability issue. This is a well-known issue with the tree setting. Given that the point-Laplace initialization did not recover the divergence factorization, it is not surprising to me that the method did not recover a tree. For the GBCD method, the point-Laplace fit after backfitting is able to recover the divergence factorization (the greedy point-Laplace fit alone does not recover the factorization).

# Further Exploration
## point-Laplace backfit initialized from true values
To test if there is a convergence issue in the point-Laplace fit, I try running symEBcovMF backfit with point-Laplace prior initialized with the true factorization.

```{r}
n <- nrow(sim_data$YYt)
L_divergence_true <- cbind(rep(1, n), 
                           rep(c(1,-1), times = c(n/2, n/2)),
                           rep(c(1, -1, 0), times = c(n/4, n/4, n/2)),
                           rep(c(0, 1, -1), times = c(n/2, n/4, n/4)))
```

First, we initialize a symEBcovMF object using $\hat{L}_{split}$.
```{r}
symebcovmf_true_init_obj <- sym_ebcovmf_init(sim_data$YYt)

true_L_normalized <- apply(L_divergence_true, 2, function(x){x/sqrt(sum(x^2))})
symebcovmf_true_init_obj$L_pm <- true_L_normalized

symebcovmf_true_init_obj$lambda <- rep(1, ncol(L_divergence_true))

symebcovmf_true_init_obj$resid_s2 <- estimate_resid_s2(S = sim_data$YYt, 
  L = symebcovmf_true_init_obj$L_pm, 
  lambda = symebcovmf_true_init_obj$lambda, 
  n = nrow(sim_data$Y), 
  K = length(symebcovmf_true_init_obj$lambda))

symebcovmf_true_init_obj$elbo <- compute_elbo(S = sim_data$YYt, 
  L = symebcovmf_true_init_obj$L_pm, 
  lambda = symebcovmf_true_init_obj$lambda, 
  resid_s2 = symebcovmf_true_init_obj$resid_s2, 
  n = nrow(sim_data$Y), 
  K = length(symebcovmf_true_init_obj$lambda), 
  KL = rep(0, length(symebcovmf_true_init_obj$lambda)))

symebcovmf_true_init_obj$fitted_gs <- rep(list(NULL), length(symebcovmf_true_init_obj$lambda))
```

I refit the lambda values keeping the factors fixed.
```{r}
symebcovmf_true_init_obj <- refit_lambda(S = sim_data$YYt, symebcovmf_true_init_obj, maxiter = 500)
```

Now, we run the backfit with point-Laplace prior.
```{r, results = 'hide'}
symebcovmf_true_init_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_true_init_obj, ebnm_fn = ebnm::ebnm_point_laplace, backfit_maxiter = 100)
```

This is a plot of the loadings estimate.
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_true_init_backfit$L_pm %*% diag(sqrt(symebcovmf_true_init_backfit$lambda)), bal_pops)
```

This is the objective function value:
```{r}
symebcovmf_true_init_backfit$elbo
```

This is a heatmap of the estimate of the Gram matrix:
```{r}
true_init_gram_est <- tcrossprod(symebcovmf_true_init_backfit$L_pm %*% diag(sqrt(symebcovmf_true_init_backfit$lambda)))
plot_heatmap(true_init_gram_est, brks = seq(0, max(abs(true_init_gram_est)), length.out = 50))
```

This is the L2 norm of the difference, excluding diagonal entries.
```{r}
compute_L2_fit(true_init_gram_est, sim_data$YYt)
```

This is a plot of fitted vs. observed values:
```{r}
ggplot(data = NULL, aes(x = as.vector(sim_data$YYt), y = as.vector(true_init_gram_est))) + geom_point() + geom_abline(slope = 1, intercept = 0, color = 'red') + xlab('Observed Values') + ylab('Fitted Values')
```

### Observations
This obtains a higher objective function value than the fit found by greedy symEBcovMF with backfit. So perhaps this is a convergence issue? Is GBCD/EBcovMF better at finding sparse solutions?

## Greedy fit of third factor
Here, I investigate the greedy fit of the third factor.

First, we apply symEBcovMF with `Kmax = 2` and compute the residual matrix.
```{r}
symebcovmf_rank2_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm::ebnm_point_laplace, K = 2, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), sign_constraint = NULL, refit_lam = TRUE)
```

```{r}
rank2_resid_matrix <- sim_data$YYt - tcrossprod(symebcovmf_rank2_fit$L_pm %*% diag(sqrt(symebcovmf_rank2_fit$lambda)))
```

This is a plot of the loadings estimate.
```{r}
plot_loadings(symebcovmf_rank2_fit$L_pm %*% diag(sqrt(symebcovmf_rank2_fit$lambda)), pop_vec)
```

This is a heatmap of the residual matrix.
```{r}
plot_heatmap(rank2_resid_matrix, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(rank2_resid_matrix)), max(abs(rank2_resid_matrix)), length.out = 50))
```

This is a plot of the first eigenvector of the residual matrix.
```{r}
resid_eigen <- eigen(rank2_resid_matrix)
resid_eigen$values

plot(resid_eigen$vectors[,1])
```

This is a heatmap of the component that corresponds to the first eigenvector.
```{r}
plot_heatmap(tcrossprod(resid_eigen$vectors[,1]), colors_range = c('blue','gray96','red'), brks = seq(-max(abs(tcrossprod(resid_eigen$vectors[,1]))), max(abs(tcrossprod(resid_eigen$vectors[,1]))), length.out = 50))
```

### Initialize fit with true factor
Now, we try greedily fitting the third factor initialized with the true factor.
```{r}
n <- ncol(sim_data$YYt)
true_fac3 <- rep(c(0, 1, -1), times = c(n/2, n/4, n/4))
symebcovmf_fac3_fit <- sym_ebcovmf_r1_fit(sim_data$YYt, symebcovmf_rank2_fit, ebnm_fn = ebnm_point_laplace, maxiter = 100, tol = 10^(-8), v_init = true_fac3, sign_constraint = NULL)
```

This is a plot of the third factor estimate.
```{r}
plot(symebcovmf_fac3_fit$L_pm[,3], ylab = 'Third Factor')
```

This is a plot of the ELBO when optimizing the fourth factor.
```{r}
fac3_idx <- which(symebcovmf_fac3_fit$vec_elbo_full == 3)
plot(symebcovmf_fac3_fit$vec_elbo_full[(fac3_idx+1): length(symebcovmf_fac3_fit$vec_elbo_full)], xlab = 'Iter', ylab = 'ELBO')
```

This is the objective function value:
```{r}
symebcovmf_fac3_fit$elbo
```

### Initialize fit with default initialization
This is the fit of the third factor if the default initialization procedure is used.
```{r}
symebcovmf_fac3_default_init_fit <- sym_ebcovmf_r1_fit(sim_data$YYt, symebcovmf_rank2_fit, ebnm_fn = ebnm_point_laplace, maxiter = 100, tol = 10^(-8), sign_constraint = NULL)
```

This is a plot of the third factor estimate.
```{r}
plot(symebcovmf_fac3_default_init_fit$L_pm[,3], ylab = 'Third Factor')
```

This is a plot of the ELBO when optimizing the third factor.
```{r}
fac3_idx <- which(symebcovmf_fac3_default_init_fit$vec_elbo_full == 3)
plot(symebcovmf_fac3_default_init_fit$vec_elbo_full[(fac3_idx+1): length(symebcovmf_fac3_default_init_fit$vec_elbo_full)], xlab = 'Iter', ylab = 'ELBO')
```

This is the objective function value:
```{r}
symebcovmf_fac3_default_init_fit$elbo
```

This has a higher objective function value. This suggests that the greedy procedure prefers the estimate from the default initialization. However, the previous section suggests that for overall fit, the objective function prefers the sparser representation. So the greedy method may not be the way to attain the sparser representation?
