---
title: "symebcovmf_bal_nonoverlap"
author: "Annie Xie"
date: "2025-04-08"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this example, we test out symEBcovMF on balanced, star-structured data.

# Example

```{r}
library(ebnm)
library(pheatmap)
library(ggplot2)
```

```{r}
source('code/symebcovmf_functions.R')
source('code/visualization_functions.R')
```

## Data Generation

```{r}
# adapted from Jason's code
# args is a list containing pop_sizes, branch_sds, indiv_sd, n_genes, and seed
sim_star_data <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  K <- length(args$pop_sizes)
  
  FF <- matrix(rnorm(K * p, sd = rep(args$branch_sds, each = p)), ncol = K)
  
  LL <- matrix(0, nrow = n, ncol = K)
  for (k in 1:K) {
    vec <- rep(0, K)
    vec[k] <- 1
    LL[, k] <- rep(vec, times = args$pop_sizes)
  }
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
pop_sizes <- rep(40, 4)
n_genes <- 1000
branch_sds <- rep(2,4)
indiv_sd <- 1
seed <- 1
sim_args = list(pop_sizes = pop_sizes, branch_sds = branch_sds, indiv_sd = indiv_sd, n_genes = n_genes, seed = seed)
sim_data <- sim_star_data(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(sim_data$LL, pop_vec)
```

# symEBcovMF
```{r}
symebcovmf_bal_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_point_exponential, K = 5, maxiter = 100, rank_one_tol = 10^(-8), tol = 10^(-8))
```

## Progression of ELBO
```{r}
symebcovmf_bal_full_elbo_vec <- symebcovmf_bal_fit$vec_elbo_full[!(symebcovmf_bal_fit$vec_elbo_full %in% c(1:length(symebcovmf_bal_fit$vec_elbo_K)))]
ggplot() + geom_line(data = NULL, aes(x = 1:length(symebcovmf_bal_full_elbo_vec), y = symebcovmf_bal_full_elbo_vec)) + xlab('Iter') + ylab('ELBO')
```


## Visualization of Estimate
This is a scatter plot of $\hat{L}$, the estimate from symEBcovMF:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_bal_fit$L_pm %*% diag(sqrt(symebcovmf_bal_fit$lambda)), bal_pops)
```

This is the objective function value attained:
```{r}
symebcovmf_bal_fit$elbo
```

## Visualization of Fit

This is a heatmap of $\hat{L}\hat{\Lambda}\hat{L}'$:
```{r}
symebcovmf_bal_fitted_vals <- tcrossprod(symebcovmf_bal_fit$L_pm %*% diag(sqrt(symebcovmf_bal_fit$lambda)))
plot_heatmap(symebcovmf_bal_fitted_vals, brks = seq(0, max(symebcovmf_bal_fitted_vals), length.out = 50))
```

This is a scatter plot of fitted values vs. observed values for the off-diagonal entries:
```{r}
diag_idx <- seq(1, prod(dim(sim_data$YYt)), length.out = ncol(sim_data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(sim_data$YYt))), diag_idx) 

ggplot(data = NULL, aes(x = c(sim_data$YYt)[off_diag_idx], y = c(symebcovmf_bal_fitted_vals)[off_diag_idx])) + geom_point() + ylim(-1, 15) + xlim(-1,15) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Observations
Similar to EBMFcov, symEBcovMF adds an intercept-like factor as the first factor. Therefore, when `Kmax = 4`, the method finds one intercept like factor and three population-effect factors. When `Kmax = 5`, the method recovers the last population effect factor.

# symEBcovMF with refit step

```{r}
symebcovmf_bal_refit_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_point_exponential, K = 5, maxiter = 100, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

## Progression of ELBO
```{r}
symebcovmf_bal_refit_full_elbo_vec <- symebcovmf_bal_refit_fit$vec_elbo_full[!(symebcovmf_bal_refit_fit$vec_elbo_full %in% c(1:length(symebcovmf_bal_refit_fit$vec_elbo_K)))]
ggplot() + geom_line(data = NULL, aes(x = 1:length(symebcovmf_bal_refit_full_elbo_vec), y = symebcovmf_bal_refit_full_elbo_vec)) + xlab('Iter') + ylab('ELBO')
```
A note: I don't think I save the ELBO value after the refitting step in vec_elbo_full. But the refitting does change this vector since it changes the residual matrix that is used when you add a new vector.

## Visualization of Estimate
This is a scatter plot of $\hat{L}_{refit}$, the estimate from symEBcovMF:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_bal_refit_fit$L_pm %*% diag(sqrt(symebcovmf_bal_refit_fit$lambda)), bal_pops)
```

This is the objective function value attained:
```{r}
symebcovmf_bal_refit_fit$elbo
```

## Visualization of Fit

This is a heatmap of $\hat{L}_{refit}\hat{\Lambda}_{refit}\hat{L}_{refit}'$:
```{r}
symebcovmf_bal_refit_fitted_vals <- tcrossprod(symebcovmf_bal_refit_fit$L_pm %*% diag(sqrt(symebcovmf_bal_refit_fit$lambda)))
plot_heatmap(symebcovmf_bal_refit_fitted_vals, brks = seq(0, max(symebcovmf_bal_refit_fitted_vals), length.out = 50))
```

This is a scatter plot of fitted values vs. observed values for the off-diagonal entries:
```{r}
diag_idx <- seq(1, prod(dim(sim_data$YYt)), length.out = ncol(sim_data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(sim_data$YYt))), diag_idx) 

ggplot(data = NULL, aes(x = c(sim_data$YYt)[off_diag_idx], y = c(symebcovmf_bal_refit_fitted_vals)[off_diag_idx])) + geom_point() + ylim(-1, 15) + xlim(-1,15) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Observations
We see that symEBcovMF with the refitting step improves upon the fit. Now, the first factor, which is similar to an intercept factor, has lower weight. Would it be possible for lambda to become zero without changing the vector? Perhaps if the residual matrix had multiple negative eigenvalues?
