---
title: "symebcovmf_gb_point_exp_backfit"
author: "Annie Xie"
date: "2025-05-07"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, I want to test out initializing symEBcovMF with a binary prior with a point-exponential symEBcovMF fit. 

## Motivation
In the tree setting, I found that symEBcovMF with the point-exponential prior did a better job at recovering tree structure than symEBcovMF with the generalized binary prior. Matthew and I hypothesized that the point-exponential prior is better at finding sparse solutions. So we want to investigate a two-step procedure which finds a fit with point-exponential prior in the first step, and then refines the fit with the generalized binary (or other binary) prior in the second step. In this analysis, I investigate a backfitting procedure -- first, I generate a fit with a point-exponential prior, and then use that as an initialization for a backfit with a binary prior.

# Packages and Functions

```{r}
library(ebnm)
library(pheatmap)
library(ggplot2)
library(symebcovmf)
```

```{r}
source('code/visualization_functions.R')
```

# Data Generation
To test this procedure, I will apply it to the tree-structured dataset. When testing out symEBcovMF, I found that the estimates from the two priors in the tree setting had the largest difference.

```{r}
sim_4pops <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(7 * p, sd = rep(args$branch_sds, each = p)), ncol = 7)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  LL <- matrix(0, nrow = n, ncol = 7)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 1, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 0, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(1, 0, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 1, 0, 0), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 1, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args = list(pop_sizes = rep(40, 4), n_genes = 1000, branch_sds = rep(2,7), indiv_sd = 1, seed = 1)
sim_data <- sim_4pops(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(sim_data$LL, pop_vec)
```

This is a plot of the eigenvalues of the Gram matrix:
```{r}
S_eigen <- eigen(sim_data$YYt)
plot(S_eigen$values) + abline(a = 0, b = 0, col = 'red')
```

This is the minimum eigenvalue:
```{r}
min(S_eigen$values)
```

# Generalized binary

## Generalized binary symEBcovMF

For comparison, we first try running greedy symEBcovMF with generalized binary prior.
```{r}
symebcovmf_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm::ebnm_generalized_binary, K = 7, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}$, the estimate from symEBcovMF:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_fit$L_pm %*% diag(sqrt(symebcovmf_fit$lambda)), bal_pops)
```

This is the objective function value attained:
```{r}
symebcovmf_fit$elbo
```

## Generalized binary symEBcovMF backfit initialized with point-exponential fit
For this procedure, I start by running symEBcovMF with a point-exponential prior with Kmax set to the inputted Kmax value. From this, I get an estimate for $L$, which I call $\hat{L}_{exp}$. Then, I backfit with a generalized-binary prior.

```{r}
symebcovmf_pexp_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm::ebnm_point_exponential, K = 7, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}_{exp}$, the estimate from point-exponential symEBcovMF. This is the initialization for the backfit:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_pexp_fit$L_pm %*% diag(sqrt(symebcovmf_pexp_fit$lambda)), bal_pops)
```

Now, we run the backfit.
```{r, results = 'hide'}
symebcovmf_pexp_fit_init <- symebcovmf_pexp_fit
symebcovmf_pexp_fit_init$fitted_gs <- rep(list(NULL), length(symebcovmf_pexp_fit$lambda))

symebcovmf_gb_fit_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_pexp_fit_init, ebnm_fn = ebnm_generalized_binary, backfit_maxiter = 100)
```

This is a scatter plot of $\hat{L}_{gb-backfit}$, the estimate from the generalized binary backfit:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_gb_fit_backfit$L_pm %*% diag(sqrt(symebcovmf_gb_fit_backfit$lambda)), bal_pops)
```

This is the objective function value attained:
```{r}
symebcovmf_gb_fit_backfit$elbo
```

## Observations

The estimate from the backfit procedure looks closer to the tree-structured loadings matrix than the estimate from greedy generalized binary symEBcovMF. Furthermore, the backfit procedure attains a much higher objective function value. This suggests that the method does prefer the tree-structured loadings matrix, and the greedy generalized binary symEBcovMF method has a convergence issue. The second factor corresponding to a subtype effect could be more binary. This could potentially be fixed by using a more strict binary prior. Similarly, the first factor corresponding to the intercept could be more binary. Another observation is the single group effect factors all have a couple of small, non-zero loading values. I'm not exactly sure why those appear. 

# Binormal
In this section, I try applying the same method with the binormal prior instead of the generalized binary prior.

```{r}
dbinormal = function (x,s,s0,lambda,log=TRUE){
  pi0 = 0.5
  pi1 = 0.5
  s2 = s^2
  s02 = s0^2
  l0 = dnorm(x,0,sqrt(lambda^2 * s02 + s2),log=TRUE)
  l1 = dnorm(x,lambda,sqrt(lambda^2 * s02 + s2),log=TRUE)
  logsum = log(pi0*exp(l0) + pi1*exp(l1))
 
  m = pmax(l0,l1)
  logsum = m + log(pi0*exp(l0-m) + pi1*exp(l1-m))
  if (log) return(sum(logsum))
  else return(exp(sum(logsum)))
}
```

```{r}
ebnm_binormal = function(x,s, g_init = NULL, fix_g = FALSE, output = ebnm_output_default(), control = NULL){
  # Add g_init to make the method run
  if(is.null(dim(x)) == FALSE){
    x <- c(x)
  }
  s0 = 0.01
  lambda = optimize(function(lambda){-dbinormal(x,s,s0,lambda,log=TRUE)},
              lower = 0, upper = max(x))$minimum
  g = ashr::normalmix(pi=c(0.5,0.5), mean=c(0,lambda), sd=c(lambda * s0,lambda * s0))
  postmean = ashr::postmean(g,ashr::set_data(x,s))
  postsd = ashr::postsd(g,ashr::set_data(x,s))
  log_likelihood <- ashr::calc_loglik(g, ashr::set_data(x,s))
  return(list(fitted_g = g, posterior = data.frame(mean=postmean,sd=postsd), log_likelihood = log_likelihood))
}
```

## Binormal symEBcovMF

For comparison, I try running regular symEBcovMF with the binormal prior.
```{r}
symebcovmf_binormal_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_binormal, K = 7, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}_{binormal}$, the estimate from binormal symEBcovMF:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_binormal_fit$L_pm %*% diag(sqrt(symebcovmf_binormal_fit$lambda)), bal_pops)
```

This is the objective function value attained:
```{r}
symebcovmf_binormal_fit$elbo
```

## Binormal symEBcovMF backfit with point-exponential initialization
Now we try the backfit procedure with the binormal prior.

```{r, results = 'hide'}
symebcovmf_binormal_fit_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_pexp_fit_init, ebnm_fn = ebnm_binormal, backfit_maxiter = 100)
```

This is a scatter plot of $\hat{L}_{binormal-backfit}$, the estimate from binormal backfit:
```{r}
bal_pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(symebcovmf_binormal_fit_backfit$L_pm %*% diag(sqrt(symebcovmf_binormal_fit_backfit$lambda)), bal_pops)
```

This is the objective function attained:
```{r}
symebcovmf_binormal_fit_backfit$elbo
```

## Observations
Similar to generalized binary symEBcovMF, the binormal symEBcovMF method fails to find all four single group effect factors. Instead, three of the factors combine two groups. The backfit procedure does a better job at recovering the tree structure in the data. Furthermore, the estimate from the backfit procedure attains a higher objective function value than the estimate from the greedy procedure. Comparing the binormal prior with the generalized binary prior, we see that the factors generated from the latter look more binary. Though this could be due to the settings of the priors. I'm guessing it would be possible to get a more binary estimate from the generalized binary prior by changing the scale parameter.
