---
title: "unbal_nonoverlap_backfit"
author: "Annie Xie"
date: "2025-05-09"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, I explore backfitting in the unbalanced non-overlapping setting.

## Motivation
When applying greedy symEBcovMF with point-exponential prior to the unbalanced non-overlapping setting, the method did a relatively good job. However, the third factor it found was not a single group effect factor. For the third factor, the most prominent effect was the third group effect. But there was also a small non-zero loading on the first group effect. In my exploration, I found that the greedy method does prefer this solution (this solution yields a higher objective function). Therefore, I want to test if backfitting will change this factor to a single group effect factor. I hypothesize that it will since the first group effect is already captured in the fourth factor.

# Packages and Functions

```{r}
devtools::load_all('~/Documents/Github/ebnm')
library(pheatmap)
library(ggplot2)
library(symebcovmf)
```

```{r}
source('code/visualization_functions.R')
```

# Data Generation

```{r}
# adapted from Jason's code
# args is a list containing pop_sizes, branch_sds, indiv_sd, n_genes, and seed
sim_star_data <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  K <- length(args$pop_sizes)
  
  FF <- matrix(rnorm(K * p, sd = rep(args$branch_sds, each = p)), ncol = K)
  
  LL <- matrix(0, nrow = n, ncol = K)
  for (k in 1:K) {
    vec <- rep(0, K)
    vec[k] <- 1
    LL[, k] <- rep(vec, times = args$pop_sizes)
  }
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
pop_sizes <- c(20,50,30,60)
n_genes <- 1000
branch_sds <- rep(2,4)
indiv_sd <- 1
seed <- 1
sim_args = list(pop_sizes = pop_sizes, branch_sds = branch_sds, indiv_sd = indiv_sd, n_genes = n_genes, seed = seed)
sim_data <- sim_star_data(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- rep(c('A','B','C','D'), times = pop_sizes)
plot_loadings(sim_data$LL, pop_vec)
```

# Regular symEBcovMF 

First, we apply greedy symEBcovMF with point-exponential prior to the Gram matrix:
```{r}
symebcovmf_unbal_refit_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_point_exponential, K = 4, maxiter = 100, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}$, the estimate from symEBcovMF:
```{r}
plot_loadings(symebcovmf_unbal_refit_fit$L_pm %*% diag(sqrt(symebcovmf_unbal_refit_fit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_unbal_refit_fit$elbo
```

# symEBcovMF with backfit

Now, we try backfitting (also with a point-exponential prior):
```{r}
symebcovmf_fit_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_unbal_refit_fit, ebnm_fn = ebnm::ebnm_point_exponential, backfit_maxiter = 100)
```

This is a scatter plot of $\hat{L}_{backfit}$, the estimate from symEBcovMF with backfit:
```{r}
plot_loadings(symebcovmf_fit_backfit$L_pm %*% diag(sqrt(symebcovmf_fit_backfit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_fit_backfit$elbo
```

# Observations

We see that after backfitting, the third factor looks closer to a single group effect factor. The first group effect that was originally present has shrunk closer to zero. Furthermore, the objective function value after the backfit was quite a bit higher. One interesting observation is that some of the values that were previously zero are now small, non-zero values. Some of the other methods that work in the covariance space also have this characteristic, e.g. GBCD, EBMFcov. So perhaps this is related to working in the covariance space?
