---
title: "unbaltree_setting_pt_exp_prior_exploration"
author: "Annie Xie"
date: "2026-02-05"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, I want to explore the symEBcovMF method in the unbalanced tree setting.

# Packages and Functions

```{r}
devtools::load_all('~/Documents/Github/ebnm')
library(pheatmap)
library(ggplot2)
library(symebcovmf)
```

```{r}
source('code/visualization_functions.R')
```

```{r}
compute_L2_fit <- function(est, dat){
  score <- sum((dat - est)^2) - sum((diag(dat) - diag(est))^2)
  return(score)
}
```

```{r}
compute_crossprod_similarity <- function(est, truth){
  K_est <- ncol(est)
  K_truth <- ncol(truth)
  n <- nrow(est)
  
  #if estimates don't have same number of columns, try padding the estimate with zeros and make cosine similarity zero
  if (K_est < K_truth){
    est <- cbind(est, matrix(rep(0, n*(K_truth-K_est)), nrow = n))
  }
  
  if (K_est > K_truth){
    truth <- cbind(truth, matrix(rep(0, n*(K_est - K_truth)), nrow = n))
  }
  
  #normalize est and truth
  norms_est <- apply(est, 2, function(x){sqrt(sum(x^2))})
  norms_est[norms_est == 0] <- Inf
  
  norms_truth <- apply(truth, 2, function(x){sqrt(sum(x^2))})
  norms_truth[norms_truth == 0] <- Inf
  
  est_normalized <- t(t(est)/norms_est)
  truth_normalized <- t(t(truth)/norms_truth)
  
  #compute matrix of cosine similarities
  cosine_sim_matrix <- abs(crossprod(est_normalized, truth_normalized))
  
  assignment_problem <- lpSolve::lp.assign(t(cosine_sim_matrix), direction = "max")
  return((1/K_truth)*assignment_problem$objval)
}
```

# Data Generation

```{r}
sim_4pops_unbal_tree <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(7 * p, sd = rep(args$branch_sds, each = p)), ncol = 7)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  # check if this is right!
  LL <- matrix(0, nrow = n, ncol = 7)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 0, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 1, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(0, 1, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 0, 1, 1), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 1, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args = list(pop_sizes = rep(40, 4), n_genes = 1000, branch_sds = rep(2,7), indiv_sd = 1, seed = 1)
pop_vec <- rep(c('A','B','C','D'), each = 40)
sim_data <- sim_4pops_unbal_tree(sim_args)
```

This is a heatmap of the true loadings matrix:
```{r}
plot_heatmap(sim_data$LL)
```

# symEBcovMF with point-exponential

First, we start with running greedy symEBcovMF with the point-exponential prior.
```{r, results = 'hide', warning = FALSE}
symebcovmf_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm::ebnm_point_exponential, K = 7, maxiter = 500, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a scatter plot of $\hat{L}_{pt-exp}$, the estimate from symEBcovMF:
```{r}
plot_loadings(symebcovmf_fit$L_pm %*% diag(sqrt(symebcovmf_fit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_fit$elbo
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit$L_pm, sim_data$LL)
```

## Observations

The representation looks more akin to a balanced tree representation compared to the unbalanced loadings matrix used to generate the data. The last two factors group two populations together, illustrating a bifurcating tree structure. The bifurcating tree structure might be a result of the non-binary baseline. This is a plot of the first factor and the number of branches each individual is a member of. 

```{r}
plot(symebcovmf_fit$L_pm %*% diag(sqrt(symebcovmf_fit$lambda))[,1], rowSums(sim_data$LL), xlab = 'first factor', ylab = 'number of branches')
```

We can see that the first factor is correlated with the number of branches each individual is a member of. I'm curious to see if backfitting would allow the representation to look more binary. However, if I am backfitting with the point-exponential prior, I don't suspect that it would. If I am backfitting with the generalized binary prior, then I think it should. 

# Additional Backfit

Now, we run additional backfit (also with the point-exponential prior). This is the code for the backfit:
```{r, results = 'hide', warning = FALSE}
symebcovmf_fit_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_fit, ebnm_fn = ebnm_point_exponential, backfit_maxiter = 1000)
```

This is a scatter plot of $\hat{L}_{pt-exp-backfit}$, the estimate from symEBcovMF with backfit:
```{r}
plot_loadings(symebcovmf_fit_backfit$L_pm %*% diag(sqrt(symebcovmf_fit_backfit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_fit_backfit$elbo
```

This is a plot of the progression of the ELBO:
```{r}
plot(symebcovmf_fit_backfit$backfit_iter_elbo_vec, ylab = 'ELBO')
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit_backfit$L_pm, sim_data$LL)
```

## Observations

Backfitting with the exponential prior did not change the representation much. This result is not surprising to me since the point-exponential prior doesn't discriminate against non-binary solutions.

# More Backfit

I also try additional backfit with the point-exponential prior to see if that changes anything.
```{r, results = 'hide', warning = FALSE}
symebcovmf_fit_longer_backfit <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_fit_backfit, ebnm_fn = ebnm_point_exponential, backfit_maxiter = 10000)
```

This is a scatter plot of $\hat{L}_{pt-exp-longer-backfit}$, the estimate from symEBcovMF with backfit:
```{r}
plot_loadings(symebcovmf_fit_longer_backfit$L_pm %*% diag(sqrt(symebcovmf_fit_longer_backfit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_fit_longer_backfit$elbo
```

This is a plot of the progression of the ELBO:
```{r}
plot(symebcovmf_fit_longer_backfit$backfit_iter_elbo_vec, ylab = 'ELBO')
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit_longer_backfit$L_pm, sim_data$LL)
```

## Observations

The baseline factor is still correlated with the number of branches each group is a member of. We see some shrinkage in the last factor which captures group 1 and group 2 effects. However, the second factor, which previously only captured group 1, has a relatively small positive loading for group 2 now. So maybe some of the shrinkage that occurred in the last factor was just the effect moving to the second factor.

# Backfit with binary prior

In this section, I try backfitting with the generalized binary prior instead of the point-exponential prior.
```{r, results = 'hide', warning = FALSE}
symebcovmf_fit_backfit_binary_prior <- symebcovmf_fit_backfit
symebcovmf_fit_backfit_binary_prior$fitted_gs <- rep(list(NULL), length(symebcovmf_fit_backfit$lambda))
symebcovmf_fit_backfit_binary_prior <- sym_ebcovmf_backfit(sim_data$YYt, symebcovmf_fit_backfit_binary_prior, ebnm_fn = ebnm::ebnm_generalized_binary, backfit_maxiter = 5000)
```

This is a scatter plot of $\hat{L}_{pt-exp-binary-backfit}$, the estimate from symEBcovMF with backfit:
```{r}
plot_loadings(symebcovmf_fit_backfit_binary_prior$L_pm %*% diag(sqrt(symebcovmf_fit_backfit_binary_prior$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_fit_backfit_binary_prior$elbo
```

This is a plot of the progression of the ELBO:
```{r}
plot(symebcovmf_fit_backfit_binary_prior$backfit_iter_elbo_vec, ylab = 'ELBO')
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_fit_backfit_binary_prior$L_pm, sim_data$LL)
```

## Observations

The representation still contains the bifurcating tree structure. So this particular prior did not help make the representation more binary.

# Fit Initialized with True Loadings

```{r}
init_L_norms <- apply(sim_data$LL, 2, function(x){sqrt(sum(x^2))})
init_L_normalized <- t(t(sim_data$LL)/init_L_norms)
init_lambda <- init_L_norms^2 * diag(crossprod(sim_data$FF)/ncol(sim_data$Y))
```

```{r, results = 'hide', warning = FALSE}
symebcovmf_true_init_obj <- sym_ebcovmf_init(sim_data$YYt)
symebcovmf_true_init_obj <- sym_ebcovmf_factors_init(sim_data$YYt, 
                                                  symebcovmf_true_init_obj, 
                                                  init_L_normalized, 
                                                  init_lambda)
  
symebcovmf_true_init_fit <- refit_lambda(S = sim_data$YYt, 
                                 symebcovmf_true_init_obj, 
                                 maxiter = 500)
symebcovmf_true_init_fit <- sym_ebcovmf_backfit(sim_data$YYt, 
                                        symebcovmf_true_init_fit, 
                                        ebnm_fn = ebnm::ebnm_generalized_binary, 
                                        backfit_maxiter = 500)
```

This is a scatter plot of $\hat{L}_{pt-exp-binary-backfit}$, the estimate from symEBcovMF with backfit:
```{r}
plot_loadings(symebcovmf_true_init_fit$L_pm %*% diag(sqrt(symebcovmf_true_init_fit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_true_init_fit$elbo
```

This is a plot of the progression of the ELBO:
```{r}
plot(symebcovmf_true_init_fit$backfit_iter_elbo_vec, ylab = 'ELBO')
```

This is the crossproduct similarity value:
```{r}
compute_crossprod_similarity(symebcovmf_true_init_fit$L_pm, sim_data$LL)
```

## Observations

The fit initialized with the true loadings matrix attains a higher ELBO than the fit initialized with the greedy fit. So it's not that the method prefers the more balanced fit. I wonder if a bottom-up style initialization would be more likely to find the unbalanced representation.
