---
title: "symebcovmf_unbal_nonoverlap"
author: "Annie Xie"
date: "2025-04-29"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this example, we test out symEBcovMF on unbalanced, star-structured data.

# Example

```{r}
library(ebnm)
library(pheatmap)
library(ggplot2)
```

```{r}
source('code/symebcovmf_functions.R')
source('code/visualization_functions.R')
```

## Data Generation

```{r}
# adapted from Jason's code
# args is a list containing pop_sizes, branch_sds, indiv_sd, n_genes, and seed
sim_star_data <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  K <- length(args$pop_sizes)
  
  FF <- matrix(rnorm(K * p, sd = rep(args$branch_sds, each = p)), ncol = K)
  
  LL <- matrix(0, nrow = n, ncol = K)
  for (k in 1:K) {
    vec <- rep(0, K)
    vec[k] <- 1
    LL[, k] <- rep(vec, times = args$pop_sizes)
  }
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
pop_sizes <- c(20,50,30,60)
n_genes <- 1000
branch_sds <- rep(2,4)
indiv_sd <- 1
seed <- 1
sim_args = list(pop_sizes = pop_sizes, branch_sds = branch_sds, indiv_sd = indiv_sd, n_genes = n_genes, seed = seed)
sim_data <- sim_star_data(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- rep(c('A','B','C','D'), times = pop_sizes)
plot_loadings(sim_data$LL, pop_vec)
```

# symEBcovMF 
```{r}
symebcovmf_unbal_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_point_exponential, K = 4, maxiter = 100, rank_one_tol = 10^(-8), tol = 10^(-8))
```

## Progression of ELBO
```{r}
symebcovmf_unbal_full_elbo_vec <- symebcovmf_unbal_fit$vec_elbo_full[!(symebcovmf_unbal_fit$vec_elbo_full %in% c(1:length(symebcovmf_unbal_fit$vec_elbo_K)))]
ggplot() + geom_line(data = NULL, aes(x = 1:length(symebcovmf_unbal_full_elbo_vec), y = symebcovmf_unbal_full_elbo_vec)) + xlab('Iter') + ylab('ELBO')
```


## Visualization of Estimate
This is a scatter plot of $\hat{L}$, the estimate from symEBcovMF:
```{r}
plot_loadings(symebcovmf_unbal_fit$L_pm %*% diag(sqrt(symebcovmf_unbal_fit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_unbal_fit$elbo
```

## Visualization of Fit

This is a heatmap of $\hat{L}\hat{\Lambda}\hat{L}'$:
```{r}
symebcovmf_unbal_fitted_vals <- tcrossprod(symebcovmf_unbal_fit$L_pm %*% diag(sqrt(symebcovmf_unbal_fit$lambda)))
plot_heatmap(symebcovmf_unbal_fitted_vals, brks = seq(0, max(symebcovmf_unbal_fitted_vals), length.out = 50))
```

This is a scatter plot of fitted values vs. observed values for the off-diagonal entries:
```{r}
diag_idx <- seq(1, prod(dim(sim_data$YYt)), length.out = ncol(sim_data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(sim_data$YYt))), diag_idx) 

ggplot(data = NULL, aes(x = c(sim_data$YYt)[off_diag_idx], y = c(symebcovmf_unbal_fitted_vals)[off_diag_idx])) + geom_point() + ylim(-1, 5) + xlim(-1,5) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Observations
symEBcovMF does a relatively good job at recovering the four population effects. Interestingly, while factor 3 primarily recovers the third group effect, it has a small (but still notable) loading on the first group effect. I explore this more in another analysis. I did also try symEBcovMF with generalized binary prior, and in that case, the factors primarily captured one group effect each.

# symEBcovMF with refit step

```{r}
symebcovmf_unbal_refit_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_point_exponential, K = 4, maxiter = 100, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

## Progression of ELBO
```{r}
symebcovmf_unbal_refit_full_elbo_vec <- symebcovmf_unbal_refit_fit$vec_elbo_full[!(symebcovmf_unbal_refit_fit$vec_elbo_full %in% c(1:length(symebcovmf_unbal_refit_fit$vec_elbo_K)))]
ggplot() + geom_line(data = NULL, aes(x = 1:length(symebcovmf_unbal_refit_full_elbo_vec), y = symebcovmf_unbal_refit_full_elbo_vec)) + xlab('Iter') + ylab('ELBO')
```
A note: I don't think I save the ELBO value after the refitting step in vec_elbo_full. But the refitting does change this vector since it changes the residual matrix that is used when you add a new vector.

## Visualization of Estimate
This is a scatter plot of $\hat{L}_{refit}$, the estimate from symEBcovMF:
```{r}
plot_loadings(symebcovmf_unbal_refit_fit$L_pm %*% diag(sqrt(symebcovmf_unbal_refit_fit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_unbal_refit_fit$elbo
```

## Visualization of Fit

This is a heatmap of $\hat{L}_{refit}\hat{\Lambda}_{refit}\hat{L}_{refit}'$:
```{r}
symebcovmf_unbal_refit_fitted_vals <- tcrossprod(symebcovmf_unbal_refit_fit$L_pm %*% diag(sqrt(symebcovmf_unbal_refit_fit$lambda)))
plot_heatmap(symebcovmf_unbal_refit_fitted_vals, brks = seq(0, max(symebcovmf_unbal_refit_fitted_vals), length.out = 50))
```

This is a scatter plot of fitted values vs. observed values for the off-diagonal entries:
```{r}
diag_idx <- seq(1, prod(dim(sim_data$YYt)), length.out = ncol(sim_data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(sim_data$YYt))), diag_idx) 

ggplot(data = NULL, aes(x = c(sim_data$YYt)[off_diag_idx], y = c(symebcovmf_unbal_refit_fitted_vals)[off_diag_idx])) + geom_point() + ylim(-1, 5) + xlim(-1,5) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Comparison of Estimates

Comparison of the factors:
```{r}
for (i in 1:4){
  print(ggplot(data = NULL, aes(x = symebcovmf_unbal_fit$L_pm[,i], y = symebcovmf_unbal_refit_fit$L_pm[,i])) + geom_point() + geom_abline(slope = 1, intercept = 0, color = 'red') + ylab('symEBcovMF with refitting') + xlab('Regular symEBcovMF') + labs(title=paste('Factor',i)))
}
```

## Observations
The estimate from symEBcovMF with refitting qualitatively looks the same as the estimate from regular symEBcovMF. Plotting the values of the two estimates show that they are nearly the same.
