---
title: "unbal_nonoverlap_exploration"
author: "Annie Xie"
date: "2025-04-29"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, I explore the unbalanced non-overlapping setting.

# Packages and Functions

```{r}
library(ebnm)
library(pheatmap)
library(ggplot2)
```

```{r}
source('code/symebcovmf_functions.R')
source('code/visualization_functions.R')
```

# Data Generation

```{r}
# adapted from Jason's code
# args is a list containing pop_sizes, branch_sds, indiv_sd, n_genes, and seed
sim_star_data <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  K <- length(args$pop_sizes)
  
  FF <- matrix(rnorm(K * p, sd = rep(args$branch_sds, each = p)), ncol = K)
  
  LL <- matrix(0, nrow = n, ncol = K)
  for (k in 1:K) {
    vec <- rep(0, K)
    vec[k] <- 1
    LL[, k] <- rep(vec, times = args$pop_sizes)
  }
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
pop_sizes <- c(20,50,30,60)
n_genes <- 1000
branch_sds <- rep(2,4)
indiv_sd <- 1
seed <- 1
sim_args = list(pop_sizes = pop_sizes, branch_sds = branch_sds, indiv_sd = indiv_sd, n_genes = n_genes, seed = seed)
sim_data <- sim_star_data(sim_args)
```

This is a heatmap of the scaled Gram matrix:
```{r}
plot_heatmap(sim_data$YYt, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(sim_data$YYt)), max(abs(sim_data$YYt)), length.out = 50))
```

This is a scatter plot of the true loadings matrix:
```{r}
pop_vec <- rep(c('A','B','C','D'), times = pop_sizes)
plot_loadings(sim_data$LL, pop_vec)
```

# symEBcovMF with refitting

```{r}
symebcovmf_unbal_refit_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_point_exponential, K = 4, maxiter = 100, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

## Visualization of Estimate
This is a scatter plot of $\hat{L}_{refit}$, the estimate from symEBcovMF:
```{r}
plot_loadings(symebcovmf_unbal_refit_fit$L_pm %*% diag(sqrt(symebcovmf_unbal_refit_fit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_unbal_refit_fit$elbo
```

## Visualization of Fit

This is a heatmap of $\hat{L}_{refit}\hat{\Lambda}_{refit}\hat{L}_{refit}'$:
```{r}
symebcovmf_unbal_refit_fitted_vals <- tcrossprod(symebcovmf_unbal_refit_fit$L_pm %*% diag(sqrt(symebcovmf_unbal_refit_fit$lambda)))
plot_heatmap(symebcovmf_unbal_refit_fitted_vals, brks = seq(0, max(symebcovmf_unbal_refit_fitted_vals), length.out = 50))
```

This is a scatter plot of fitted values vs. observed values for the off-diagonal entries:
```{r}
diag_idx <- seq(1, prod(dim(sim_data$YYt)), length.out = ncol(sim_data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(sim_data$YYt))), diag_idx) 

ggplot(data = NULL, aes(x = c(sim_data$YYt)[off_diag_idx], y = c(symebcovmf_unbal_refit_fitted_vals)[off_diag_idx])) + geom_point() + ylim(-1, 5) + xlim(-1,5) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Observations
As noted in a different analysis, symEBcovMF does a relatively good job at recovering the four population effects. However, one undesired aspect of the estimate is that factor 3 contains the effects of two different groups. I would like to figure out why this is happening. Furthermore, the group 1 effect is much smaller than the group 4 effect, so I'm wondering if it's possible to shrink the group 1 effect down to zero so that the factor only captures one group.

# Trying a smaller convergence tolerance
In this section, we explore whether a smaller convergence tolerance will yield a factor which captures only one group. Perhaps the group 1 effect is in the process of being (slowly) shrunk down to zero, and a smaller tolerance will allow for more shrinkage.

```{r}
symebcovmf_unbal_refit_smaller_tol_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_point_exponential, K = 4, maxiter = 100, rank_one_tol = 10^(-15), tol = 10^(-8), refit_lam = TRUE)
```

## Visualization of Estimate
This is a scatter plot of $\hat{L}_{refit}$, the estimate from symEBcovMF:
```{r}
plot_loadings(symebcovmf_unbal_refit_smaller_tol_fit$L_pm %*% diag(sqrt(symebcovmf_unbal_refit_smaller_tol_fit$lambda)), pop_vec)
```

This is the objective function value attained:
```{r}
symebcovmf_unbal_refit_smaller_tol_fit$elbo
```

Comparison of factor 3 estimates:
```{r}
ggplot(data = NULL, aes(x = symebcovmf_unbal_refit_fit$L_pm[,3], y = symebcovmf_unbal_refit_smaller_tol_fit$L_pm[,3])) + geom_point() + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Observations
Decreasing the convergence tolerance does not lead to more shrinkage of the group 1 effect in the factor 3 estimate. The optimization procedure for factor 3 does stop because the ELBO slightly decreases (on the order of $10^{-11}$). My guess is the decrease is caused by a numerical issue. So it's possible that we would see more shrinkage if we let the optimization run for more iterations.

# Try initializing from true factor
To check if this is a convergence issue, I try fitting the third factor initialized from the true single population effect factor. 

First, we fit the first two factors.
```{r}
symebcovmf_unbal_refit_rank2_fit <- sym_ebcovmf_fit(S = sim_data$YYt, ebnm_fn = ebnm_point_exponential, K = 2, maxiter = 100, rank_one_tol = 10^(-8), tol = 10^(-8), refit_lam = TRUE)
```

This is a heatmap of the residual matrix, $S - \sum_{k=1}^{2} \hat{\lambda}_k \hat{\ell}_k \hat{\ell}_k'$:
```{r}
R <- sim_data$YYt - tcrossprod(symebcovmf_unbal_refit_rank2_fit$L_pm %*% diag(sqrt(symebcovmf_unbal_refit_rank2_fit$lambda)))
plot_heatmap(R, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(R)), max(abs(R)), length.out = 50))
```

Now, we fit the third factor, initialized with the true population effect factor.
```{r}
symebcovmf_unbal_true_init_fac3_fit <- sym_ebcovmf_r1_fit(S = sim_data$YYt, symebcovmf_unbal_refit_rank2_fit, ebnm_fn = ebnm::ebnm_point_exponential, maxiter = 100, tol = 10^(-8), v_init = rep(c(0,0,1,0), times = pop_sizes))
```

This is a plot of the estimate of the third factor.
```{r}
plot(symebcovmf_unbal_true_init_fac3_fit$L_pm[,3], ylab = 'Third Factor')
```

This is a plot of the ELBO during the fit of the third factor.
```{r}
fac3.idx <- which(symebcovmf_unbal_true_init_fac3_fit$vec_elbo_full == 3)
plot(symebcovmf_unbal_true_init_fac3_fit$vec_elbo_full[(fac3.idx+1):length(symebcovmf_unbal_true_init_fac3_fit$vec_elbo_full)], ylab = 'ELBO')
```

```{r}
true_fac3 <- rep(c(0,0,1,0), times = pop_sizes)
true_fac3 <- true_fac3/sqrt(sum(true_fac3^2))
estimates_list <- list(true_fac3)
for (i in 1:11){
  estimates_list[[(i+1)]] <- sym_ebcovmf_r1_fit(sim_data$YYt, symebcovmf_unbal_refit_rank2_fit, ebnm_fn = ebnm::ebnm_point_exponential, maxiter = i, tol = 10^(-8), v_init = rep(c(0,0,1,0), times = pop_sizes))$L_pm[,3]
}
```

This is a plot of the progression of the estimate.
```{r}
par(mfrow = c(6,2), mar = c(2, 2, 1, 1) + 0.1)
max_y <- max(sapply(estimates_list, max))
min_y <- min(sapply(estimates_list, min))
for (i in 1:12){
  plot(estimates_list[[i]], main = paste('Iter', (i-1)), ylab = 'L', ylim = c(min_y, max_y))
}
par(mfrow = c(1,1))
```

## Observations
When initialized with the true factor, the method still yields an estimate that has non-zero loading on the first group. This suggests that the method does prefer this estimate. Perhaps an easier way of getting a representation comprised of four single group effect factors is to backfit. Question I have: why does this method have this issue but EBMFcov-greedy does not?

